---
title: "Mofa_Integration"
author: "Ali Raza, Pritom, niloufer"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

MOFA2 multi-omics integration and molecular subtyping. Picks up the aligned
RNA/methylation/mutation matrices from 01_QC_EDA.Rmd, learns 15 latent
factors, then clusters patients on those factors to get candidate subtypes.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

data_dir <- "../data"
out_dir  <- "../Outputs"
fig_dir  <- file.path(out_dir, "02_mofa_integration")   # this file's plots go here
dir.create(data_dir, showWarnings = FALSE)
dir.create(out_dir,  showWarnings = FALSE)
dir.create(fig_dir,  showWarnings = FALSE)

library(MOFA2)
library(ggplot2)
library(dplyr)

# the aligned, analysis-ready layers built in 01_QC_EDA.Rmd

rna_final  <- readRDS(file.path(data_dir, "rna_final.rds"))
meth_final <- readRDS(file.path(data_dir, "meth_final.rds"))
mut_final  <- readRDS(file.path(data_dir, "mut_final.rds"))
clin_final <- readRDS(file.path(data_dir, "clin_final.rds"))


cat("data loaded\n")
```

## MOFA2 setup

MOFA2 (Multi-Omics Factor Analysis) learns a shared latent factor space
across the three omics layers - each factor picks up variation that shows
up consistently across modalities, which is what makes it more informative
than looking at RNA, methylation or mutations on their own. We use the top
5,000 variable features for RNA and methylation, all 100 mutation genes,
15 factors, and a Gaussian/Gaussian/Bernoulli likelihood setup matching the
continuous vs. binary nature of each layer.

```{r mofa-input}
# most signal lives in the high-variance features, so subset down to those
top_rna_idx  <- order(apply(rna_final,  1, var), decreasing = TRUE)[1:5000]
top_meth_idx <- order(apply(meth_final, 1, var), decreasing = TRUE)[1:5000]

# MOFA2 wants one matrix per view, features x samples
data_mofa <- list(
  RNA         = rna_final[top_rna_idx,  ],
  Methylation = meth_final[top_meth_idx, ],
  Mutations   = mut_final
)

cat("MOFA2 input ready:\n")
cat("RNA:        ", dim(data_mofa$RNA),         "\n")
cat("Methylation:", dim(data_mofa$Methylation), "\n")
cat("Mutations:  ", dim(data_mofa$Mutations),   "\n")

mofa <- create_mofa(data_mofa)
print(mofa)
```

Set the number of factors and the per-view likelihood, then hand it off to
`prepare_mofa()` which does some internal bookkeeping before training can
start.

```{r mofa-options}
model_opts             <- get_default_model_options(mofa)
model_opts$num_factors <- 15
model_opts$likelihoods <- c(RNA         = "gaussian",
                             Methylation = "gaussian",
                             Mutations   = "bernoulli")

train_opts      <- get_default_training_options(mofa)
train_opts$seed <- 42

mofa <- prepare_mofa(mofa,
                     model_options    = model_opts,
                     training_options = train_opts)
cat("MOFA2 ready to train\n")
```

Actual training happens here - this is the slow, one-time step, so it's
switched off by default and only needs flipping to `eval=TRUE` once.

```{r mofa-run, eval=FALSE}
# takes 10-30 minutes, only needs to run once - eval=FALSE keeps it off
# during normal knits, flip to TRUE to (re)train
mofa <- run_mofa(mofa, use_basilisk = TRUE)
saveRDS(mofa, file.path(data_dir, "mofa_model.rds"))
cat("MOFA2 trained and saved\n")
```

Every other run just reloads the already-trained model from disk instead
of retraining.

```{r mofa-reload}
# training already happened above (or in a previous run) - just reload it
mofa <- readRDS(file.path(data_dir, "mofa_model.rds"))
```

## Variance explained per factor

Factors that explain variance in more than one layer are the ones that
actually reflect shared multi-omics signal, rather than noise specific to
one layer.

```{r mofa-variance}
p1 <- plot_variance_explained(mofa, max_r2 = 15)
p1
ggsave(file.path(fig_dir, "variance_explained_per_factor.png"), p1, width = 8, height = 6)

# total variance captured per layer
p2 <- plot_variance_explained(mofa, plot_total = TRUE)[[2]]
p2
ggsave(file.path(fig_dir, "variance_explained_total.png"), p2, width = 6, height = 5)
```

Pull the actual factor scores (one value per patient per factor) out of
the trained model - this patient x factor matrix is what the rest of the
pipeline (clustering, survival) actually works with.

```{r extract-factors}
factors <- get_factors(mofa, factors = "all")[[1]]
cat("factor matrix dimensions:", dim(factors), "\n")
```

## Molecular subtyping

k-means on the factor matrix, with k chosen by silhouette score rather than
picked arbitrarily.

```{r silhouette-scores}
library(cluster)

sil_scores <- sapply(2:6, function(k) {
  set.seed(42)
  km <- kmeans(factors, centers = k, nstart = 25)
  mean(silhouette(km$cluster, dist(factors))[, 3])
})

best_k <- (2:6)[which.max(sil_scores)]
cat("best k:", best_k, "\n")

sil_df <- data.frame(k = 2:6, silhouette = sil_scores)
p <- ggplot(sil_df, aes(k, silhouette)) +
  geom_line(color = "#2E4A7A") +
  geom_point(size = 3, color = "#2E4A7A") +
  geom_vline(xintercept = best_k, linetype = "dashed", color = "#C0392B") +
  theme_minimal() +
  labs(title = "Silhouette Score by Number of Clusters",
       x = "Number of clusters (k)", y = "Mean silhouette width")
p
ggsave(file.path(fig_dir, "silhouette_scores.png"), p, width = 7, height = 5)
```

Refit k-means at the best k from above to get the actual subtype labels.

```{r kmeans-final}
set.seed(42)

km_final       <- kmeans(factors, centers = best_k, nstart = 25)
subtype_labels <- km_final$cluster

cat("subtype distribution:\n")
print(table(subtype_labels))
```

Visualize the subtypes on a UMAP of the factor matrix, to see whether the
k-means clusters correspond to visually distinct groups.

```{r umap-subtypes}
library(umap)

umap_out <- umap(factors)
df_umap  <- data.frame(
  UMAP1   = umap_out$layout[, 1],
  UMAP2   = umap_out$layout[, 2],
  Subtype = factor(subtype_labels)
)

p <- ggplot(df_umap, aes(UMAP1, UMAP2, color = Subtype)) +
  geom_point(size = 1.5, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Molecular Subtypes - UMAP of MOFA2 Factors",
       color = "Subtype")
p
ggsave(file.path(fig_dir, "umap_subtypes.png"), p, width = 7, height = 5)
```

## Survival association per MOFA2 factor

Split each factor at its median (high vs low) and run a log-rank test
against overall survival - a quick way to screen all 15 factors before
committing to the ones that actually track survival.

```{r factor-survival}
library(survival)

clin_m <- clin_final[match(substr(rownames(factors), 1, 12),
                            clin_final$submitter_id), ]
time   <- ifelse(clin_m$vital_status == "Dead",
                 clin_m$days_to_death, clin_m$days_to_last_follow_up)
event  <- ifelse(clin_m$vital_status == "Dead", 1, 0)

pval_df <- data.frame(
  Factor  = paste0("Factor", 1:15),
  p_value = sapply(1:15, function(i) {
    d  <- data.frame(time, event, g = ifelse(factors[,i] > median(factors[,i]), "High", "Low"))
    d  <- d[!is.na(d$time) & d$time > 0, ]
    lr <- tryCatch(survdiff(Surv(time, event) ~ g, data = d), error = function(e) NULL)
    if (is.null(lr)) return(NA)
    1 - pchisq(lr$chisq, df = 1)
  })
) |> arrange(p_value)

print(pval_df)

p <- ggplot(pval_df, aes(x = reorder(Factor, p_value), y = p_value)) +
  geom_bar(stat = "identity", fill = "#2E4A7A", alpha = 0.8) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "#C0392B") +
  coord_flip() + theme_minimal() +
  labs(title = "Survival Association per MOFA2 Factor",
       x = "Factor", y = "Log-rank p-value") +
  annotate("text", x = 1, y = 0.06, label = "p = 0.05", color = "#C0392B", size = 3)
p
ggsave(file.path(fig_dir, "factor_survival_pvalues.png"), p, width = 7, height = 6)
```

## Factor weight inspection

Factor 12 and Factor 6 are the ones that turn out to matter for survival
(next section) - look at which genes and mutations actually drive them
before using them further.

```{r check-weights}
# quick look at which features drive the two factors we care about later
weights_f12_rna <- get_weights(mofa, views = "RNA",       factors = 12)[[1]]
weights_f12_mut <- get_weights(mofa, views = "Mutations", factors = 12)[[1]]
weights_f6_rna  <- get_weights(mofa, views = "RNA",       factors = 6)[[1]]

cat("Factor 12 RNA - top 10 weights:\n")
print(round(sort(abs(weights_f12_rna[,1]), decreasing=TRUE)[1:10], 4))

cat("\nFactor 12 Mutations - top 10 weights:\n")
print(round(sort(abs(weights_f12_mut[,1]), decreasing=TRUE)[1:10], 4))

cat("\nFactor 6 RNA - top 10 weights:\n")
print(round(sort(abs(weights_f6_rna[,1]), decreasing=TRUE)[1:10], 4))
```

## Export ML feature matrix (Factor 6 & 12 top genes)

Factors 6 and 12 are carried into the survival analysis next - pull their
top genes and build the feature matrix the Cox model will use.

```{r export-ml-features}
get_top <- function(factor_num, n = 25) {
  w <- get_weights(mofa, views = "RNA", factors = factor_num)[[1]]
  data.frame(gene = rownames(w), weight = w[,1]) |>
    arrange(desc(abs(weight))) |> head(n)
}

top_genes <- rbind(get_top(12), get_top(6)) |>
             distinct(gene, .keep_all = TRUE)

cat("total unique genes:", nrow(top_genes), "\n")

# keep only patients that also have clinical data, then build the feature matrix
common_pts <- intersect(substr(colnames(rna_final), 1, 12), clin_final$submitter_id)
rna_sub    <- rna_final[rownames(rna_final) %in% top_genes$gene,
                         substr(colnames(rna_final), 1, 12) %in% common_pts]

ml_df           <- as.data.frame(t(rna_sub))
colnames(ml_df) <- make.names(colnames(ml_df))

cat("patients:", nrow(ml_df), "| features:", ncol(ml_df), "\n")
write.csv(ml_df, file.path(data_dir, "top_extracted50_genes.csv"), row.names = TRUE)
cat("saved\n")
```
