---
title: "Survival_Analysis"
author: "Ali Raza, Pritom, niloufer"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

Cox proportional-hazards model on the top Factor 6 & 12 genes. Picks up
clinical data from 01_QC_EDA.Rmd and the ML feature matrix exported by
02_mofa_integration.Rmd (which also covers the per-factor survival
screening).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

data_dir <- "../data"
out_dir  <- "../Outputs"
fig_dir  <- file.path(out_dir, "04_survival_analysis")   # this file's plots go here
dir.create(data_dir, showWarnings = FALSE)
dir.create(out_dir,  showWarnings = FALSE)
dir.create(fig_dir,  showWarnings = FALSE)

library(ggplot2)
library(dplyr)
library(survival)
library(survminer)

clin_final <- readRDS(file.path(data_dir, "clin_final.rds"))

cat("data loaded\n")
```

## Cox proportional-hazards model

Cox model on the top Factor 6 & 12 genes exported from 02_mofa_integration.Rmd.

```{r cox-regression}
ml_df <- read.csv(file.path(data_dir, "top_extracted50_genes.csv"), row.names = 1)

sample_ids   <- substr(rownames(ml_df), 1, 12)
clin_matched <- clin_final[match(sample_ids, clin_final$submitter_id), ]

ml_df$time  <- ifelse(clin_matched$vital_status == "Dead",
                       clin_matched$days_to_death,
                       clin_matched$days_to_last_follow_up)
ml_df$event <- ifelse(clin_matched$vital_status == "Dead", 1, 0)
ml_df$label <- NULL

ml_df <- ml_df[!is.na(ml_df$time) & ml_df$time > 0, ]
cat("patients in Cox:", nrow(ml_df), "\n")

cox_fit <- coxph(Surv(time, event) ~ ., data = ml_df, x = TRUE)
cat("C-index:", round(cox_fit$concordance["concordance"], 3), "\n")
summary(cox_fit)

# 07_annotation.Rmd needs the fitted model to pull gene coefficients
saveRDS(cox_fit, file.path(data_dir, "cox_fit.rds"))
```

Quick sanity check on the cohort that actually went into the Cox model -
how many patients, how many events vs. censored, and the follow-up range.

```{r event-summary}
cat("total patients in Cox:    ", nrow(ml_df), "\n")
cat("total deaths (events):    ", sum(ml_df$event == 1), "\n")
cat("censored (alive/unknown): ", sum(ml_df$event == 0), "\n")
cat("median follow-up (days):  ", round(median(ml_df$time), 0), "\n")
cat("max follow-up (days):     ", max(ml_df$time), "\n")
```

Split patients into high/low risk by their Cox linear predictor (median
split) and plot Kaplan-Meier curves for the two groups - this is the plot
that actually shows whether the model separates survival outcomes.

```{r cox-km}
# linear predictor already gives us a risk score, no need to call predict()
pred <- cox_fit$linear.predictors
ml_df$risk_group <- ifelse(pred > median(pred), "High", "Low")

fit_risk <- survfit(Surv(time, event) ~ risk_group, data = ml_df)

km_plot <- ggsurvplot(fit_risk, data = ml_df, pval = TRUE,
           palette      = c("#C0392B", "#2E4A7A"),
           title        = "KM - High vs Low Cox Risk Score",
           xlab         = "Days",
           ylab         = "Survival Probability",
           legend.title = "Risk Group",
           ggtheme      = theme_minimal())
km_plot
ggsave(file.path(fig_dir, "cox_km_risk_groups.png"), km_plot$plot, width = 7, height = 5)
```

Rank genes by absolute Cox coefficient and plot them as risk-increasing or
protective - these 15 genes are what 07_annotation.Rmd digs into further.

```{r top-cox-genes-plot}
cox_summary <- as.data.frame(summary(cox_fit)$coefficients)
cox_summary$gene <- rownames(cox_summary)

top_genes_cox <- cox_summary |>
  arrange(desc(abs(coef))) |>
  dplyr::select(gene, coef, `exp(coef)`, `Pr(>|z|)`) |>
  head(15)

colnames(top_genes_cox) <- c("gene", "coef", "hazard_ratio", "p_value")
top_genes_cox$direction <- ifelse(top_genes_cox$coef > 0, "Risk", "Protective")

p <- ggplot(top_genes_cox, aes(x = reorder(gene, abs(coef)),
                           y = coef,
                           fill = direction)) +
  geom_bar(stat = "identity", width = 0.7, alpha = 0.85) +
  scale_fill_manual(values = c("Risk" = "#C0392B", "Protective" = "#2E4A7A")) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 15 Genes Driving the Cox Survival Model",
       x     = "Gene",
       y     = "Cox Coefficient",
       fill  = "Direction") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  theme(plot.title = element_text(face = "bold", size = 13))
p
ggsave(file.path(fig_dir, "top_cox_genes.png"), p, width = 7, height = 6)
```
