---
title: "QC_EDA"
author: "Ali Raza, Pritom, niloufer"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

Data acquisition, per-layer preprocessing and QC for the CRC multi-omics
project. Picks up where the raw GDC downloads left off and produces the
cleaned RNA/methylation/mutation matrices used by the MOFA2 step.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# paths are relative to this file (Analysis/), so ../data and ../Outputs
# point at the project-level folders
data_dir <- "../data"
out_dir  <- "../Outputs"
fig_dir  <- file.path(out_dir, "01_eda")   # this file's plots go here
dir.create(data_dir, showWarnings = FALSE)
dir.create(out_dir,  showWarnings = FALSE)
dir.create(fig_dir,  showWarnings = FALSE)

library(SummarizedExperiment)
library(TCGAbiolinks)
library(DESeq2)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(reshape2)
library(ggplot2)
library(dplyr)
library(MOFA2)

# these come from the download chunks below, cached so we don't redownload
rna_data  <- readRDS(file.path(data_dir, "rna_data.rds"))
meth_data <- readRDS(file.path(data_dir, "meth_data.rds"))
mut_data  <- readRDS(file.path(data_dir, "mut_data.rds"))
clinical  <- readRDS(file.path(data_dir, "clinical.rds"))

cat("data loaded\n")
```

## Data acquisition

Raw data comes from GDC via TCGAbiolinks: RNA-seq (STAR counts), methylation
(450K beta values), somatic mutations (masked MAF) and clinical metadata for
TCGA-COAD + TCGA-READ. Only needs to run once, so these chunks are switched
off (`eval=FALSE`) - flip to TRUE for a first run.

Data download: none of this is stored in the repo (too large for GitHub) -
the chunks below pull it programmatically from the GDC Data Portal:
[TCGA-COAD](https://portal.gdc.cancer.gov/projects/TCGA-COAD),
[TCGA-READ](https://portal.gdc.cancer.gov/projects/TCGA-READ).

```{r download-rna, eval=FALSE}
# raw STAR counts, not TPM - DESeq2 needs integer counts
query_rna <- GDCquery(
  project       = c("TCGA-COAD", "TCGA-READ"),
  data.category = "Transcriptome Profiling",
  data.type     = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)

GDCdownload(query_rna)
rna_data <- GDCprepare(query_rna)
```

```{r download-meth, eval=FALSE}
# Illumina 450K beta values, same two projects
query_meth <- GDCquery(
  project       = c("TCGA-COAD", "TCGA-READ"),
  data.category = "DNA Methylation",
  data.type     = "Methylation Beta Value",
  platform      = "Illumina Human Methylation 450"
)

GDCdownload(query_meth, files.per.chunk = 10)
meth_data <- GDCprepare(query_meth)
```

```{r download-mut, eval=FALSE}
# open access masked MAF, no dbGaP approval needed
query_mut <- GDCquery(
  project       = c("TCGA-COAD", "TCGA-READ"),
  data.category = "Simple Nucleotide Variation",
  data.type     = "Masked Somatic Mutation",
  access        = "open"
)

GDCdownload(query_mut)
mut_data <- GDCprepare(query_mut)
```

```{r download-clinical, eval=FALSE}
# clinical metadata isn't in a GDCquery/GDCdownload flow, it's its own endpoint
clin_coad <- GDCquery_clinic(project = "TCGA-COAD", type = "clinical")
clin_read <- GDCquery_clinic(project = "TCGA-READ",  type = "clinical")
clinical  <- bind_rows(clin_coad, clin_read)
saveRDS(clinical, file.path(data_dir, "clinical.rds"))
```

```{r save-raw, eval=FALSE}
# cache the raw objects so we never have to download GDC again
saveRDS(rna_data,  file.path(data_dir, "rna_data.rds"))
saveRDS(meth_data, file.path(data_dir, "meth_data.rds"))
saveRDS(mut_data,  file.path(data_dir, "mut_data.rds"))
```

## Raw data QC

Quick sanity check before any filtering - dimensions look right and primary
tumor samples dominate each layer.

```{r qc-dimensions}
dim(assay(rna_data))    # ~60,000 genes x ~460 samples
dim(assay(meth_data))   # ~450,000 probes x ~460 samples
dim(mut_data)

table(rna_data$sample_type)
table(meth_data$sample_type)

dim(clinical)
colnames(clinical)
```

## Per-layer preprocessing

Each layer gets cleaned separately before merging: RNA-seq is VST
normalized, methylation is filtered down to usable probes and converted to
M-values, mutations get collapsed to a binary matrix over the top mutated
genes.

### RNA-seq normalization

Keep primary tumor samples only, then pull out the raw STAR read counts -
that's the assay DESeq2 needs (not TPM or FPKM).

```{r rna-filter-primary}
# drop normals/metastases, keep primary tumor only
rna_pt <- rna_data[, rna_data$sample_type == "Primary Tumor"]

cat("available assays:\n")
print(assayNames(rna_pt))

rna_mat <- assay(rna_pt, "unstranded")
cat("RNA matrix dimensions:", dim(rna_mat), "\n")
```

Run VST (variance-stabilizing transform) on the count matrix after
dropping genes with too few reads across too few samples - this is
normalization for downstream analysis, not a differential expression test,
so the DESeq2 design is intentionally just an intercept.

```{r rna-vst}
rna_mat <- as.matrix(rna_mat)
rna_mat <- round(rna_mat)
storage.mode(rna_mat) <- "integer"

cat("total genes:        ", nrow(rna_mat), "\n")
cat("duplicate gene IDs: ", sum(duplicated(rownames(rna_mat))), "\n")

# no covariates here, just want VST normalization not DE testing
coldata <- data.frame(cond = factor(rep("tumor", ncol(rna_mat))))
rownames(coldata) <- colnames(rna_mat)

cat("matrix columns:", ncol(rna_mat), "| coldata rows:", nrow(coldata), "\n")

dds <- DESeqDataSetFromMatrix(
  countData = rna_mat,
  colData   = coldata,
  design    = ~ 1
)
dds <- dds[rowSums(counts(dds) >= 10) >= 10, ]
cat("after low-count filter:", nrow(dds), "genes\n")

# blind=TRUE since we're normalizing, not testing a design
vsd      <- vst(dds, blind = TRUE)
rna_norm <- assay(vsd)

cat("RNA normalized:", dim(rna_norm), "\n")
```

### Methylation cleaning

Keep primary tumor samples, drop probes that are unreliable for one reason
or another (missing values, sex chromosomes, SNP overlap, known
cross-reactivity), then convert beta values to M-values for downstream
statistics.

```{r meth-filter-probes}
meth_pt  <- meth_data[, meth_data$sample_type == "Primary Tumor"]
meth_mat <- assay(meth_pt)

# drop probes with any missing values
meth_mat <- meth_mat[complete.cases(meth_mat), ]

anno <- as.data.frame(
  getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
)

# sex chromosome probes confound with patient sex, drop them
sex_probes <- rownames(anno)[anno$chr %in% c("chrX", "chrY")]
meth_mat   <- meth_mat[!rownames(meth_mat) %in% sex_probes, ]

# probes overlapping a SNP pick up genotype, not methylation
snp_probes <- rownames(anno)[!is.na(anno$Probe_rs) & anno$Probe_rs != ""]
meth_mat   <- meth_mat[!rownames(meth_mat) %in% snp_probes, ]

# known cross-reactive 450K probes (Chen et al. list), cache locally
cross_probes_file <- file.path(data_dir, "cross_reactive_probes.txt")
if (!file.exists(cross_probes_file)) {
  download.file(
    "https://raw.githubusercontent.com/sirselim/illumina450k_filtering/master/48639-non-specific-probes-Illumina450k.csv",
    cross_probes_file
  )
}
cross_probes <- read.csv(cross_probes_file)$TargetID
meth_mat     <- meth_mat[!rownames(meth_mat) %in% cross_probes, ]

# M-values behave better statistically than betas (less compressed near 0/1)
# clip first so log() doesn't blow up at the extremes
meth_mat[meth_mat <= 0.001] <- 0.001
meth_mat[meth_mat >= 0.999] <- 0.999
meth_M <- log2(meth_mat / (1 - meth_mat))

cat("methylation cleaned:", dim(meth_M), "\n")
```

### Mutation binary matrix

Take the 100 most frequently mutated genes and turn the mutation calls
into a simple patient x gene matrix of 0/1 (mutated or not) - MOFA2's
Bernoulli likelihood expects binary input like this.

```{r mut-binary-matrix}
library(reshape2)

top_genes_mut <- names(sort(table(mut_data$Hugo_Symbol), decreasing = TRUE))[1:100]
mut_simple    <- mut_data[mut_data$Hugo_Symbol %in% top_genes_mut,
                           c("Tumor_Sample_Barcode", "Hugo_Symbol")]

mut_matrix <- reshape2::dcast(
  mut_simple,
  Hugo_Symbol ~ Tumor_Sample_Barcode,
  fun.aggregate = function(x) as.integer(length(x) > 0),
  fill = 0
)
rownames(mut_matrix) <- mut_matrix$Hugo_Symbol
mut_matrix <- as.matrix(mut_matrix[, -1])

cat("mutation matrix dimensions:", dim(mut_matrix), "\n")
```

### Lining up patients across layers

MOFA2 needs the same patients, in the same column order, in every matrix.
Match on the first 12 characters of the TCGA barcode (that's the patient
ID - the rest is sample/portion/plate info) and keep only patients present
in all three layers.

```{r align-patients-initial}
rna_patients  <- substr(colnames(rna_norm),   1, 12)
meth_patients <- substr(colnames(meth_M),     1, 12)
mut_patients  <- substr(colnames(mut_matrix), 1, 12)

common_patients <- Reduce(intersect, list(rna_patients, meth_patients, mut_patients))
cat("patients present in all three layers:", length(common_patients), "\n")

rna_final  <- rna_norm[,   match(common_patients, rna_patients)]
meth_final <- meth_M[,     match(common_patients, meth_patients)]
mut_final  <- mut_matrix[, match(common_patients, mut_patients)]

# use plain patient IDs as column names now that everything lines up
colnames(rna_final) <- colnames(meth_final) <- colnames(mut_final) <- common_patients

# mutations aren't touched by ComBat, so this version is already final
saveRDS(mut_final, file.path(data_dir, "mut_final.rds"))
```

## Exploratory QC plots

Before integration: check for batch effects in RNA/methylation (PCA by
tissue source site), look at the mutation landscape, and see how the
samples separate on UMAP.

### Batch effect check - RNA PCA

PCA on the most variable genes, colored by tissue source site (TSS) - if
samples cluster by TSS rather than biology, that's a batch effect we need
to correct before integration.

```{r pca-rna-batch}
# TSS = tissue source site, positions 6-7 of the TCGA barcode, stands in for batch
tss <- substr(colnames(rna_final), 6, 7)
cat("TSS codes in RNA data:\n")
print(table(tss))

top_var <- order(apply(rna_final, 1, var), decreasing = TRUE)[1:500]
pca     <- prcomp(t(rna_final[top_var, ]), scale. = TRUE)

df <- data.frame(
  PC1   = pca$x[, 1],
  PC2   = pca$x[, 2],
  batch = tss
)

p <- ggplot(df, aes(PC1, PC2, color = batch)) +
  geom_point(size = 1.5, alpha = 0.7) +
  theme_minimal() +
  labs(title = "RNA-seq PCA - colored by TSS")
p
ggsave(file.path(fig_dir, "rna_pca_batch.png"), p, width = 7, height = 5)
```

### Batch effect check - methylation PCA

Same check on methylation, using the most variable CpG probes instead of
genes.

```{r pca-meth-batch}
tss_meth <- substr(colnames(meth_final), 6, 7)

top_cpg <- order(apply(meth_final, 1, var), decreasing = TRUE)[1:10000]
pca_m   <- prcomp(t(meth_final[top_cpg, ]), scale. = TRUE)

df_m <- data.frame(
  PC1   = pca_m$x[, 1],
  PC2   = pca_m$x[, 2],
  batch = tss_meth
)

p <- ggplot(df_m, aes(PC1, PC2, color = batch)) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Methylation PCA by batch (TSS)")
p
ggsave(file.path(fig_dir, "meth_pca_batch.png"), p, width = 7, height = 5)
```

The PCA plots above are a visual check - this turns it into a number by
regressing each principal component on TSS and looking at the R2, so we
have something concrete to compare before vs. after correction.

```{r batch-quantitative}
# how much of PC1/PC2 does batch explain, R2 from a simple lm
tss_rna  <- substr(colnames(rna_norm), 6, 7)
pca_rna  <- prcomp(t(rna_norm[order(apply(rna_norm, 1, var),
                                     decreasing=TRUE)[1:500], ]), scale.=TRUE)
r2_rna_pc1 <- summary(lm(pca_rna$x[,1] ~ factor(tss_rna)))$r.squared
r2_rna_pc2 <- summary(lm(pca_rna$x[,2] ~ factor(tss_rna)))$r.squared

tss_meth <- substr(colnames(meth_M), 6, 7)
pca_meth <- prcomp(t(meth_M[order(apply(meth_M, 1, var),
                                    decreasing=TRUE)[1:1000], ]), scale.=TRUE)
r2_meth_pc1 <- summary(lm(pca_meth$x[,1] ~ factor(tss_meth)))$r.squared
r2_meth_pc2 <- summary(lm(pca_meth$x[,2] ~ factor(tss_meth)))$r.squared

batch_df <- data.frame(
  Layer = c("RNA", "RNA", "Methylation", "Methylation"),
  PC    = c("PC1", "PC2", "PC1", "PC2"),
  R2    = round(c(r2_rna_pc1, r2_rna_pc2, r2_meth_pc1, r2_meth_pc2), 3)
)
print(batch_df)
cat("\nrule of thumb: R2 < 0.10 negligible, 0.10-0.30 moderate, > 0.30 strong batch effect\n")
```

R2 above 0.1 is enough to justify correcting for it, so run ComBat on both
RNA and methylation, then rebuild the patient-aligned matrices from the
corrected data - these are the final versions that get saved and used
from here on.

```{r combat-correction}
library(sva)

# correct RNA for TSS batch, overwrite rna_norm
batch_rna <- substr(colnames(rna_norm), 6, 7)
rna_norm  <- ComBat(dat = rna_norm, batch = batch_rna,
                    mod = NULL, par.prior = TRUE)
cat("RNA ComBat done:", dim(rna_norm), "\n")

# same for methylation
batch_meth <- substr(colnames(meth_M), 6, 7)
meth_M     <- ComBat(dat = meth_M, batch = batch_meth,
                     mod = NULL, par.prior = TRUE)
cat("methylation ComBat done:", dim(meth_M), "\n")

# rebuild the aligned matrices from the corrected data and save the final versions
rna_final  <- rna_norm[, match(common_patients, substr(colnames(rna_norm), 1, 12))]
meth_final <- meth_M[,   match(common_patients, substr(colnames(meth_M),   1, 12))]
colnames(rna_final)  <- common_patients
colnames(meth_final) <- common_patients

saveRDS(rna_final,  file.path(data_dir, "rna_final.rds"))
saveRDS(meth_final, file.path(data_dir, "meth_final.rds"))
```

Recheck the same R2 metric after correction, for RNA first.

```{r rna-batcheffect}
# recheck batch effect after ComBat - should be much lower
tss_rna <- substr(colnames(rna_norm), 6, 7)
pca_rna <- prcomp(t(rna_norm[order(apply(rna_norm, 1, var),
                                    decreasing=TRUE)[1:500], ]), scale.=TRUE)
r2_pc1 <- summary(lm(pca_rna$x[,1] ~ factor(tss_rna)))$r.squared
r2_pc2 <- summary(lm(pca_rna$x[,2] ~ factor(tss_rna)))$r.squared

cat("RNA after ComBat:\n")
cat("R2 PC1:", round(r2_pc1, 3), "\n")
cat("R2 PC2:", round(r2_pc2, 3), "\n")
cat("\nbefore ComBat: PC1=0.235, PC2=0.166\n")
```

Same check for methylation.

```{r batch-after-combat-meth}
tss_meth <- substr(colnames(meth_M), 6, 7)
pca_meth <- prcomp(t(meth_M[order(apply(meth_M, 1, var),
                                    decreasing=TRUE)[1:1000], ]), scale.=TRUE)
r2_meth_pc1 <- summary(lm(pca_meth$x[,1] ~ factor(tss_meth)))$r.squared
r2_meth_pc2 <- summary(lm(pca_meth$x[,2] ~ factor(tss_meth)))$r.squared

cat("methylation after ComBat:\n")
cat("R2 PC1:", round(r2_meth_pc1, 3), "\n")
cat("R2 PC2:", round(r2_meth_pc2, 3), "\n")
cat("\nbefore ComBat: PC1=0.167, PC2=0.118\n")
```

Build a simple binary survival label (died within 3 years vs. survived
past 3 years) purely for coloring the UMAP plots below - the Cox model
later on uses actual time-to-event data, not this label.

```{r clinical-labels}
clin_final <- readRDS(file.path(data_dir, "clinical.rds"))

# binary 3-year survival label, only used for EDA/UMAP coloring
# the Cox model later on does not use this label but yes, we will run binary model as well 
clin_final$label <- NA
clin_final$label[clin_final$vital_status == "Dead" &
                 clin_final$days_to_death <= 1095] <- 0
clin_final$label[clin_final$days_to_last_follow_up > 1095] <- 1

saveRDS(clin_final, file.path(data_dir, "clin_final.rds"))
cat("label distribution (0=died<3yr, 1=survived>3yr):\n")
print(table(clin_final$label, useNA = "always"))
```

### UMAP - RNA-seq colored by survival label

UMAP on the most variable genes (post-ComBat), colored by the survival
label - a quick look at whether survival separates out in the RNA data
even before running MOFA2.

```{r umap-rna}
library(umap)
label_lookup <- setNames(clin_final$label, clin_final$submitter_id)
top_var      <- order(apply(rna_final, 1, var), decreasing = TRUE)[1:2000]
um_rna       <- umap(t(rna_final[top_var, ]))

p <- ggplot(data.frame(
    UMAP1    = um_rna$layout[,1],
    UMAP2    = um_rna$layout[,2],
    survival = label_lookup[substr(colnames(rna_final), 1, 12)]),
  aes(UMAP1, UMAP2, color = factor(survival))) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_color_manual(values   = c("0" = "#C0392B", "1" = "#2E4A7A"),
                     labels   = c("0" = "Did not survive", "1" = "Survived"),
                     na.value = "grey80") +
  theme_minimal() +
  labs(title = "RNA-seq UMAP - survival label", color = "Survival")
p
ggsave(file.path(fig_dir, "umap_rna_survival.png"), p, width = 7, height = 5)
```

### UMAP - methylation colored by survival label

Same check on the methylation layer.

```{r umap-meth}
top_cpg  <- order(apply(meth_final, 1, var), decreasing = TRUE)[1:1000]
um_meth  <- umap(t(meth_final[top_cpg, ]))

p <- ggplot(data.frame(
    UMAP1    = um_meth$layout[, 1],
    UMAP2    = um_meth$layout[, 2],
    survival = label_lookup[substr(colnames(meth_final), 1, 12)]),
  aes(UMAP1, UMAP2, color = factor(survival))) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_color_manual(values = c("0" = "#C0392B", "1" = "#2E4A7A"),
                     labels = c("0" = "Did not survive", "1" = "Survived"),
                     na.value = "grey80") +
  theme_minimal() +
  labs(title = "Methylation UMAP - survival label", color = "Survival")
p
ggsave(file.path(fig_dir, "umap_meth_survival.png"), p, width = 7, height = 5)
```

### OncoPrint - top 20 mutated genes

Standard mutation landscape plot - which genes are hit, how often, and
whether patients share the same mutations or not.

```{r oncoprint}
library(maftools)

maf_obj <- read.maf(maf = mut_data)

grDevices::png(file.path(fig_dir, "oncoprint.png"), width = 1400, height = 900, res = 120)
oncoplot(maf = maf_obj, top = 20, fontSize = 12)
title("Top 20 mutated genes - CRC cohort")
grDevices::dev.off()

# also show it inline in the knitted report
oncoplot(maf = maf_obj, top = 20)
```

### Tumor mutation burden (TMB) distribution

Total mutation count per patient, on a log scale - CRC typically shows a
bimodal split between MSI-H (hypermutated) and MSS tumors.

```{r tmb-histogram}
# TMB is bimodal in CRC 
tmb <- maf_obj@variants.per.sample

p <- ggplot(tmb, aes(x = Variants)) +
  geom_histogram(bins = 50, fill = "#2E4A7A", color = "white", alpha = 0.8) +
  scale_x_log10() +
  theme_minimal() +
  labs(
    title = "Tumor Mutation Burden - CRC cohort",
    x     = "Number of mutations per patient (log scale)",
    y     = "Number of patients"
  )
p
ggsave(file.path(fig_dir, "tmb_histogram.png"), p, width = 7, height = 5)
```

## Cross-modal patient alignment

Take the patients present in all three omics layers and confirm the column
order matches across matrices - MOFA2 needs that for the next step.

```{r align-patients}
cat("RNA final:  ", dim(rna_final),  "\n")
cat("meth final: ", dim(meth_final), "\n")
cat("mut final:  ", dim(mut_final),  "\n")

cat("RNA vs meth aligned:", all(colnames(rna_final) == colnames(meth_final)), "\n")
cat("RNA vs mut aligned: ", all(colnames(rna_final) == colnames(mut_final)),  "\n")
```
