---
title: "External_scRNA_Validation"
author: "Ali Raza, Pritom, niloufer"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

External validation of the Factor 6 & 12 top genes (from
02_mofa_integration.Rmd) against an independent single-cell CRC dataset
(GSE132465). The Cox model in 04_survival_analysis.Rmd was trained on bulk
RNA-seq, which averages over every cell in a tumor sample - checking the
same genes in single-cell data tells us whether they're actually driven by
a specific cell type, rather than being a bulk-only artifact.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

data_dir <- "../data"
out_dir  <- "../Outputs"
fig_dir  <- file.path(out_dir, "03_scrna_validation")   # this file's plots go here
gse_dir  <- file.path(data_dir, "GSE132465")         # raw scRNA-seq download
dir.create(data_dir, showWarnings = FALSE)
dir.create(out_dir,  showWarnings = FALSE)
dir.create(fig_dir,  showWarnings = FALSE)

library(ggplot2)

cat("data loaded\n")
```

## Load the subsampled Seurat object

`sc_small.rds` is a 500-cells-per-type subsample already pulled out of the
full ~1GB GSE132465 matrix (see `build-sc-small` below for how it was
built).

```{r scrna-load-cached}
library(Seurat)

sc <- readRDS(file.path(data_dir, "sc_small.rds"))
cat("loaded:", ncol(sc), "cells x", nrow(sc), "genes\n")
```

This is how `sc_small.rds` itself was produced - kept here for reference,
switched off since the cached file already exists.

```{r build-sc-small, eval=FALSE}
library(data.table)

meta <- read.table(
  gzfile(file.path(gse_dir, "GSE132465_GEO_processed_CRC_10X_cell_annotation.txt.gz")),
  header = TRUE, row.names = 1, sep = "\t"
)
cat("total cells:", nrow(meta), "\n")

set.seed(42)

# mast cells only have 3 tumor cells here, not enough to say anything useful
tumor_meta <- meta[meta$Class == "Tumor" & meta$Cell_type != "Mast cells", ]

sampled_cells <- do.call(rbind, lapply(
  unique(tumor_meta$Cell_type), function(ct) {
    cells <- rownames(tumor_meta[tumor_meta$Cell_type == ct, ])
    n     <- min(500, length(cells))
    data.frame(barcode = sample(cells, n), Cell_type = ct)
  }
))

matrix_file <- file.path(gse_dir, "GSE132465_GEO_processed_CRC_10X_natural_log_TPM_matrix.txt.gz")

# peek at the header first so we only read the columns we actually want
header    <- fread(matrix_file, nrows = 0)
all_cols  <- names(header)
keep_cols <- c(all_cols[1], sampled_cells$barcode[sampled_cells$barcode %in% all_cols])

mat_small <- fread(matrix_file, select = keep_cols)
mat <- as.matrix(mat_small[, -1])
rownames(mat) <- mat_small[[1]]

meta_small <- meta[colnames(mat), ]
sc <- CreateSeuratObject(counts = mat, meta.data = meta_small)
saveRDS(sc, file.path(data_dir, "sc_small.rds"))
```

## QC and normalize

Standard scRNA-seq QC - drop cells with too few or too many detected genes
(empty droplets / doublets) and cells with high mitochondrial content
(dying cells), then normalize so expression is comparable across cells.

```{r scrna-qc-normalize}
sc[["percent.mt"]] <- PercentageFeatureSet(sc, pattern = "^MT-")
sc <- subset(sc, subset = nFeature_RNA > 200 &
                          nFeature_RNA < 6000 &
                          percent.mt   < 25)
cat("after QC:", ncol(sc), "cells\n")

sc <- NormalizeData(sc)
```

## Map genes to symbols and check overlap

The top genes from 02_mofa_integration.Rmd are ENSEMBL IDs (with version
numbers); scRNA-seq gene names are plain symbols, so convert and see how
many of our genes are even present in this dataset.

```{r convert-and-overlap}
library(org.Hs.eg.db)

top_genes_f6  <- colnames(read.csv(file.path(data_dir, "top_extracted50_genes.csv"), row.names = 1))
ensembl_clean <- gsub("\\..*", "", top_genes_f6)

gene_symbols <- mapIds(
  org.Hs.eg.db,
  keys      = ensembl_clean,
  column    = "SYMBOL",
  keytype   = "ENSEMBL",
  multiVals = "first"
)
gene_symbols <- gene_symbols[!is.na(gene_symbols)]
cat("successfully converted:", length(gene_symbols), "genes\n")
print(gene_symbols)

overlap <- intersect(gene_symbols, rownames(sc))
cat("genes found in scRNA-seq:", length(overlap), "out of", length(gene_symbols), "\n")
```

## Expression heatmap by cell type

Average expression of the overlapping genes within each cell type - a
gene that's specific to one cell type will show up as a bright single
column here rather than being spread evenly across all of them.

```{r factor6-heatmap}
library(tidyr)

Idents(sc) <- "Cell_type"

avg_long <- AverageExpression(sc, features = overlap,
                               group.by = "Cell_type")$RNA |>
            as.data.frame() |>
            tibble::rownames_to_column("gene") |>
            pivot_longer(-gene, names_to = "cell_type",
                         values_to = "expression")

p <- ggplot(avg_long, aes(x = cell_type, y = gene, fill = expression)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_gradient(low = "lightgrey", high = "#C0392B") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
        axis.text.y = element_text(size = 9, face = "bold"),
        panel.grid  = element_blank(),
        plot.title  = element_text(face = "bold", size = 15)) +
  labs(title = "Factor 6 & 12 Genes - Expression per Cell Type",
       x = "Cell Type", y = NULL, fill = "Avg Expression")

# tall image - long gene list stacked vertically
ggsave(file.path(fig_dir, "factor6&12_heatmap.png"), plot = p, width = 12, height = 18, dpi = 200)
cat("saved: factor6_heatmap.png\n")
```
