---
title: "Biological_Annotation"
author: "Ali Raza, Pritom, niloufer"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

Biological annotation of the Cox model's top genes - readable gene names
and a pathway enrichment to interpret the biology driving the survival
model. Picks up the fitted Cox model from 04_survival_analysis.Rmd. The
scRNA-seq cell-type validation of these same genes lives in its own
file, 03_external_scrna_validate.Rmd.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

data_dir <- "../data"
out_dir  <- "../Outputs"
fig_dir  <- file.path(out_dir, "07_annotation")   # this file's plots go here
dir.create(data_dir, showWarnings = FALSE)
dir.create(out_dir,  showWarnings = FALSE)
dir.create(fig_dir,  showWarnings = FALSE)

library(ggplot2)
library(dplyr)
library(org.Hs.eg.db)
library(gridExtra)
library(grid)

cox_fit <- readRDS(file.path(data_dir, "cox_fit.rds"))

cat("data loaded\n")
```

## Gene annotation and pathway enrichment

### Annotate the top Cox genes

Pull the 15 genes with the largest absolute coefficient in the Cox model
and attach a readable symbol, full name, hazard ratio and risk direction
to each.

```{r gene-annotation}
cox_summary      <- as.data.frame(summary(cox_fit)$coefficients)
cox_summary$gene <- rownames(cox_summary)

top15 <- cox_summary |>
  arrange(desc(abs(coef))) |>
  head(15)

ensembl_clean <- gsub("\\..*", "", top15$gene)
symbols  <- mapIds(org.Hs.eg.db, keys = ensembl_clean,
                   column = "SYMBOL",   keytype = "ENSEMBL", multiVals = "first")
genename <- mapIds(org.Hs.eg.db, keys = ensembl_clean,
                   column = "GENENAME", keytype = "ENSEMBL", multiVals = "first")

annot_df <- data.frame(
  symbol       = symbols,
  full_name    = genename,
  hazard_ratio = round(top15$`exp(coef)`, 3),
  direction    = ifelse(top15$coef > 0, "Risk", "Protective")
)

print(annot_df)
write.csv(annot_df, file.path(data_dir, "top15_annotated_genes.csv"), row.names = FALSE)
cat("saved: top15_annotated_genes.csv\n")

# plain table image - just the data, bold headers, no fills
plain_theme <- ttheme_default(
  base_size = 11,
  core = list(bg_params = list(fill = "white", col = NA))
)
gene_table <- tableGrob(annot_df, rows = NULL, theme = plain_theme)

# size the image to the table's actual dimensions so nothing gets cropped
gt_w <- grid::convertWidth(sum(gene_table$widths),   "in", valueOnly = TRUE)
gt_h <- grid::convertHeight(sum(gene_table$heights), "in", valueOnly = TRUE)
ggsave(file.path(fig_dir, "annotated_genes_table.png"), gene_table,
       width = gt_w + 0.3, height = gt_h + 0.3, dpi = 150)
cat("saved: annotated_genes_table.png\n")
```

### Pathway enrichment

Run these gene symbols through g:Profiler against GO Biological Process,
KEGG and Reactome to see what pathways they're actually part of.

```{r gprofiler-enrichment}
library(gprofiler2)

gene_list <- annot_df$symbol[!is.na(annot_df$symbol)]
cat("genes for enrichment:", paste(gene_list, collapse = ", "), "\n")

gost_res <- gost(query = gene_list,
                 organism = "hsapiens",
                 sources  = c("GO:BP", "KEGG", "REAC"),
                 correction_method = "fdr")

top_paths <- gost_res$result |>
  dplyr::arrange(p_value) |>
  dplyr::select(source, term_name, p_value, intersection_size) |>
  head(15)
print(top_paths)

# styled table image - dark header + alternating row stripes, same blue
# used in the plots throughout this project
paths_display <- top_paths
paths_display$p_value <- formatC(paths_display$p_value, format = "e", digits = 2)
colnames(paths_display) <- c("Database", "Pathway", "p-value", "Genes")

striped_theme <- ttheme_default(
  base_size = 10,
  core = list(bg_params = list(fill = rep(c("white", "grey95"), length.out = nrow(paths_display)),
                                col = NA)),
  colhead = list(fg_params = list(col = "white", fontface = "bold"),
                 bg_params = list(fill = "#2E4A7A", col = NA))
)
paths_table <- tableGrob(paths_display, rows = NULL, theme = striped_theme)

# size the image to the table's actual dimensions so nothing gets cropped
pt_w <- grid::convertWidth(sum(paths_table$widths),   "in", valueOnly = TRUE)
pt_h <- grid::convertHeight(sum(paths_table$heights), "in", valueOnly = TRUE)
ggsave(file.path(fig_dir, "pathway_enrichment_table.png"), paths_table,
       width = pt_w + 0.3, height = pt_h + 0.3, dpi = 150)
cat("saved: pathway_enrichment_table.png\n")

p <- ggplot(top_paths, aes(x = reorder(term_name, -log10(p_value)),
                       y = -log10(p_value),
                       fill = source)) +
  geom_bar(stat = "identity", alpha = 0.85) +
  coord_flip() +
  scale_fill_manual(values = c("GO:BP" = "#2E4A7A",
                                "KEGG"  = "#C0392B",
                                "REAC"  = "#27AE60")) +
  theme_minimal() +
  theme(axis.text.y  = element_text(size = 9),
        plot.title   = element_text(face = "bold", size = 13),
        legend.title = element_text(size = 10)) +
  labs(title = "Pathway Enrichment - Top Cox Survival Genes",
       x     = "Pathway",
       y     = "-log10(adjusted p-value)",
       fill  = "Database")
p
ggsave(file.path(fig_dir, "gprofiler_barplot.png"), p, width = 12, height = 8, dpi = 200)
cat("saved: gprofiler_barplot.png\n")
```
