---
title: "Multi-Omics Integration in Colorectal Cancer (TCGA-COAD/READ)"
author: "Ali Raza,"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
  pdf_document:
    toc: true
---

# ─────────────────────────────────────────────────────────────
# PROJECT OVERVIEW
#
# Goal: Identify molecular subtypes of colorectal cancer (CRC)
#       using multi-omics integration (RNA-seq + methylation +
#       somatic mutations) and validate their clinical relevance.
#
# Dataset: TCGA-COAD + TCGA-READ (~386 primary tumor patients)
# Omics layers:
#   - RNA-seq       : STAR-Counts → VST normalized
#   - Methylation   : Illumina 450K → M-values (beta-to-M transform)
#   - Mutations     : Masked somatic MAF → binary gene-level matrix
#   - Clinical      : survival time, vital status, stage, age, sex
#
# Analysis pipeline:
#   1. Data acquisition via TCGAbiolinks
#   2. Per-layer preprocessing and QC
#   3. Patient-level alignment across modalities
#   4. MOFA2 multi-omics integration (15 latent factors)  ← this file ends here
#   5. k-means subtyping (silhouette-optimized k)
#   6. Kaplan-Meier / log-rank survival validation
# ─────────────────────────────────────────────────────────────


# ══════════════════════════════════════════════════════════════
# SECTION 0 — SETUP
# Load all libraries up front; reload pre-saved RDS objects so
# you never have to re-download (~hours) after the first run.
# ══════════════════════════════════════════════════════
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
setwd("C:/Users/Admin/Desktop/New folder/")

library(SummarizedExperiment)
library(TCGAbiolinks)
library(DESeq2)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(reshape2)
library(ggplot2)
library(dplyr)
library(MOFA2)

rna_data   <- readRDS("rna_data.rds")
meth_data  <- readRDS("meth_data.rds")
mut_data   <- readRDS("mut_data.rds")
clinical   <- readRDS("clinical.rds")
mut_matrix <- readRDS("mut_matrix.rds")

if (file.exists("rna_norm.rds"))   rna_norm   <- readRDS("rna_norm.rds")
if (file.exists("meth_M.rds"))     meth_M     <- readRDS("meth_M.rds")
if (file.exists("rna_final.rds"))  rna_final  <- readRDS("rna_final.rds")
if (file.exists("meth_final.rds")) meth_final <- readRDS("meth_final.rds")
if (file.exists("mut_final.rds"))  mut_final  <- readRDS("mut_final.rds")
if (file.exists("clin_final.rds")) clin_final <- readRDS("clin_final.rds")
if (file.exists("mofa_model.rds")) mofa       <- readRDS("mofa_model.rds")

cat("All data loaded\n")
```




# ══════════════════════════════════════════════════════════════
# SECTION 1 — DATA ACQUISITION
# Download all four data types from GDC via TCGAbiolinks.
# Run ONCE; all objects are saved as .rds after download.
# Expected sizes:
#   RNA-seq      : ~60,000 genes  × ~460 samples (SummarizedExperiment)
#   Methylation  : ~450,000 probes × ~460 samples (SummarizedExperiment)
#   Mutations    : ~50,000 rows × several MAF columns
#   Clinical     : ~600 rows × ~80 columns
# ══════════════════════════════════════════════════════════════

```{r download-rna, eval=FALSE}
# ── 1a. RNA-seq: STAR-Counts (raw integer counts, not TPM) ──
query_rna <- GDCquery(
  project       = c("TCGA-COAD", "TCGA-READ"),
  data.category = "Transcriptome Profiling",
  data.type     = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)

GDCdownload(query_rna)
rna_data <- GDCprepare(query_rna)   # returns a SummarizedExperiment
```

```{r download-meth, eval=FALSE}
# ── 1b. Methylation: Illumina 450K beta values ──
query_meth <- GDCquery(
  project       = c("TCGA-COAD", "TCGA-READ"),
  data.category = "DNA Methylation",
  data.type     = "Methylation Beta Value",
  platform      = "Illumina Human Methylation 450"
)

GDCdownload(query_meth, files.per.chunk = 10)
meth_data <- GDCprepare(query_meth)
```

```{r download-mut, eval=FALSE}
# ── 1c. Somatic mutations: Masked MAF (open access) ──
query_mut <- GDCquery(
  project       = c("TCGA-COAD", "TCGA-READ"),
  data.category = "Simple Nucleotide Variation",
  data.type     = "Masked Somatic Mutation",
  access        = "open"
)

GDCdownload(query_mut)
mut_data <- GDCprepare(query_mut)
```

```{r download-clinical, eval=FALSE}
# ── 1d. Clinical metadata (both projects merged) ──
clin_coad <- GDCquery_clinic(project = "TCGA-COAD", type = "clinical")
clin_read <- GDCquery_clinic(project = "TCGA-READ",  type = "clinical")
clinical  <- bind_rows(clin_coad, clin_read)
saveRDS(clinical, "clinical.rds")
```

```{r save-raw, eval=FALSE}
# ── Save raw objects so you never need to re-download ──
saveRDS(rna_data,  "rna_data.rds")
saveRDS(meth_data, "meth_data.rds")
saveRDS(mut_data,  "mut_data.rds")
```


# ══════════════════════════════════════════════════════════════
# SECTION 2 — RAW DATA QC
# Quick sanity check on dimensions and sample types before any
# filtering. Confirm you have tumor + normal labels and the
# right number of patients.
# ══════════════════════════════════════════════════════════════

```{r qc-dimensions}
# ── Dimensions ──
dim(assay(rna_data))    # expect ~60,000 genes × ~460 samples
dim(assay(meth_data))   # expect ~450,000 probes × ~460 samples
dim(mut_data)           # expect ~50,000 rows × several MAF columns

# ── Sample types: should see "Primary Tumor" as dominant ──
table(rna_data$sample_type)
table(meth_data$sample_type)

# ── Clinical overview ──
dim(clinical)
colnames(clinical)
```


# ══════════════════════════════════════════════════════════════
# SECTION 3 — PER-LAYER PREPROCESSING
# Each omics layer goes through its own cleaning pipeline before
# being merged. Key steps per layer:
#   RNA-seq     : keep Primary Tumor → VST normalization (DESeq2)
#   Methylation : keep Primary Tumor → probe filtering → beta→M
#   Mutations   : subset to top-100 mutated genes → binary matrix
#   Clinical    : create binary 3-year survival label
# ══════════════════════════════════════════════════════════════

## 3a. RNA-seq normalization

```{r rna-filter-primary}
# Filter to primary tumor samples only (drop normals / metastases)
rna_pt <- rna_data[, rna_data$sample_type == "Primary Tumor"]

# STAR-Counts stores raw reads in the "unstranded" assay
cat("Available assays:\n")
print(assayNames(rna_pt))

rna_mat <- assay(rna_pt, "unstranded")
cat("RNA matrix dimensions:", dim(rna_mat), "\n")
```

```{r rna-vst}
# ── Convert to proper integer matrix for DESeq2 ──
rna_mat <- as.matrix(rna_mat)
rna_mat <- round(rna_mat)
storage.mode(rna_mat) <- "integer"

cat("Total genes:         ", nrow(rna_mat), "\n")
cat("Duplicate gene IDs:  ", sum(duplicated(rownames(rna_mat))), "\n")

# ── Build minimal colData (no covariates; design=~1 for normalization only) ──
coldata <- data.frame(cond = factor(rep("tumor", ncol(rna_mat))))
rownames(coldata) <- colnames(rna_mat)

cat("Matrix columns:", ncol(rna_mat), "| Coldata rows:", nrow(coldata), "\n")

# ── DESeq2 object; filter genes with <10 counts in <10 samples ──
dds <- DESeqDataSetFromMatrix(
  countData = rna_mat,
  colData   = coldata,
  design    = ~ 1
)
dds <- dds[rowSums(counts(dds) >= 10) >= 10, ]
cat("After low-count filter:", nrow(dds), "genes\n")

# ── VST: variance stabilizing transform (blind=TRUE = no design info used) ──
vsd     <- vst(dds, blind = TRUE)
rna_norm <- assay(vsd)

saveRDS(rna_norm, "rna_norm.rds")
cat("RNA normalized:", dim(rna_norm), "\n")
```


## 3b. Methylation cleaning

```{r meth-filter-probes}
# Filter to primary tumor samples
meth_pt  <- meth_data[, meth_data$sample_type == "Primary Tumor"]
meth_mat <- assay(meth_pt)

# Remove probes with any NA across patients
meth_mat <- meth_mat[complete.cases(meth_mat), ]

# ── Load 450K probe annotation ──
anno <- as.data.frame(
  getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
)

# Remove sex chromosome probes (chrX / chrY) — confounded by sex
sex_probes <- rownames(anno)[anno$chr %in% c("chrX", "chrY")]
meth_mat   <- meth_mat[!rownames(meth_mat) %in% sex_probes, ]

# Remove SNP-overlapping probes (genotype artefacts)
snp_probes <- rownames(anno)[!is.na(anno$Probe_rs) & anno$Probe_rs != ""]
meth_mat   <- meth_mat[!rownames(meth_mat) %in% snp_probes, ]

# Remove known cross-reactive probes (multi-mapping to off-target loci)
if (!file.exists("cross_reactive_probes.txt")) {
  download.file(
    "https://raw.githubusercontent.com/sirselim/illumina450k_filtering/master/48639-non-specific-probes-Illumina450k.csv",
    "cross_reactive_probes.txt"
  )
}
cross_probes <- read.csv("cross_reactive_probes.txt")$TargetID
meth_mat     <- meth_mat[!rownames(meth_mat) %in% cross_probes, ]

# ── Beta → M-value transform: log2(beta / (1 - beta)) ──
# M-values are more statistically well-behaved (homoscedastic).
# Clip extreme betas first to prevent log(0) / log(Inf).
meth_mat[meth_mat <= 0.001] <- 0.001
meth_mat[meth_mat >= 0.999] <- 0.999
meth_M <- log2(meth_mat / (1 - meth_mat))

saveRDS(meth_M, "meth_M.rds")
cat("Methylation cleaned:", dim(meth_M), "\n")
```


## 3c. Mutation binary matrix
```{r mut-binary-matrix}
library(reshape2)

top_genes_mut <- names(sort(table(mut_data$Hugo_Symbol), decreasing = TRUE))[1:100]
mut_simple    <- mut_data[mut_data$Hugo_Symbol %in% top_genes_mut,
                           c("Tumor_Sample_Barcode", "Hugo_Symbol")]

# Force reshape2 dcast explicitly
mut_matrix <- reshape2::dcast(
  mut_simple,
  Hugo_Symbol ~ Tumor_Sample_Barcode,
  fun.aggregate = function(x) as.integer(length(x) > 0),
  fill = 0
)
rownames(mut_matrix) <- mut_matrix$Hugo_Symbol
mut_matrix <- as.matrix(mut_matrix[, -1])

cat("Mutation matrix dimensions:", dim(mut_matrix), "\n")
saveRDS(mut_matrix, "mut_matrix.rds")
```



# ══════════════════════════════════════════════════════════════
# SECTION 4 — EXPLORATORY QC PLOTS
# Before integration, visually verify:
#   - No major batch effects in RNA / methylation (PCA by TSS)
#   - Mutation landscape (OncoPrint, TMB distribution)
#   - UMAP structure of RNaseq+methyl omics layer
# ══════════════════════════════════════════════════════════════

## 4a. Batch effect check — RNA PCA

```{r pca-rna-batch}
# TSS = Tissue Source Site (positions 6-7 of TCGA barcode) — proxy for batch
tss <- substr(colnames(rna_final), 6, 7)
cat("TSS codes in RNA data:\n")
print(table(tss))

# PCA on top 500 most variable genes
top_var <- order(apply(rna_final, 1, var), decreasing = TRUE)[1:500]
pca     <- prcomp(t(rna_final[top_var, ]), scale. = TRUE)

df <- data.frame(
  PC1   = pca$x[, 1],
  PC2   = pca$x[, 2],
  batch = tss
)

ggplot(df, aes(PC1, PC2, color = batch)) +
  geom_point(size = 1.5, alpha = 0.7) +
  theme_minimal() +
  labs(title = "RNA-seq PCA — colored by TSS")
```

## 4b. Batch effect check — Methylation PCA

```{r pca-meth-batch}
# Same logic: extract TSS from methylation sample barcodes
tss_meth <- substr(colnames(meth_final), 6, 7)

# PCA on top 10,000 most variable CpGs
top_cpg <- order(apply(meth_final, 1, var), decreasing = TRUE)[1:10000]
pca_m   <- prcomp(t(meth_final[top_cpg, ]), scale. = TRUE)

df_m <- data.frame(
  PC1   = pca_m$x[, 1],
  PC2   = pca_m$x[, 2],
  batch = tss_meth
)

ggplot(df_m, aes(PC1, PC2, color = batch)) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Methylation PCA by batch (TSS)")
```



```{r batch-quantitative}
# RNA — how much does batch explain PC1 and PC2?
tss_rna  <- substr(colnames(rna_norm), 6, 7)
pca_rna  <- prcomp(t(rna_norm[order(apply(rna_norm, 1, var),
                                     decreasing=TRUE)[1:500], ]), scale.=TRUE)
r2_rna_pc1 <- summary(lm(pca_rna$x[,1] ~ factor(tss_rna)))$r.squared
r2_rna_pc2 <- summary(lm(pca_rna$x[,2] ~ factor(tss_rna)))$r.squared

# Methylation
tss_meth <- substr(colnames(meth_M), 6, 7)
pca_meth <- prcomp(t(meth_M[order(apply(meth_M, 1, var),
                                    decreasing=TRUE)[1:1000], ]), scale.=TRUE)
r2_meth_pc1 <- summary(lm(pca_meth$x[,1] ~ factor(tss_meth)))$r.squared
r2_meth_pc2 <- summary(lm(pca_meth$x[,2] ~ factor(tss_meth)))$r.squared

batch_df <- data.frame(
  Layer = c("RNA", "RNA", "Methylation", "Methylation"),
  PC    = c("PC1", "PC2", "PC1", "PC2"),
  R2    = round(c(r2_rna_pc1, r2_rna_pc2, r2_meth_pc1, r2_meth_pc2), 3)
)
print(batch_df)
cat("\nR² < 0.10 = negligible batch effect\n")
cat("R² 0.10-0.30 = moderate batch effect\n")
cat("R² > 0.30 = strong batch effect\n")
```
```{r combat-correction}
library(sva)

# RNA ComBat — save back to rna_norm
batch_rna <- substr(colnames(rna_norm), 6, 7)
rna_norm  <- ComBat(dat = rna_norm, batch = batch_rna,
                    mod = NULL, par.prior = TRUE)
saveRDS(rna_norm, "rna_norm.rds")
cat("RNA ComBat done:", dim(rna_norm), "\n")

# Methylation ComBat — save back to meth_M
batch_meth <- substr(colnames(meth_M), 6, 7)
meth_M     <- ComBat(dat = meth_M, batch = batch_meth,
                     mod = NULL, par.prior = TRUE)
saveRDS(meth_M, "meth_M.rds")
cat("Methylation ComBat done:", dim(meth_M), "\n")
```
```{r rna-batcheffect}
# Check batch effect AFTER ComBat
tss_rna <- substr(colnames(rna_norm), 6, 7)
pca_rna <- prcomp(t(rna_norm[order(apply(rna_norm, 1, var),
                                    decreasing=TRUE)[1:500], ]), scale.=TRUE)
r2_pc1 <- summary(lm(pca_rna$x[,1] ~ factor(tss_rna)))$r.squared
r2_pc2 <- summary(lm(pca_rna$x[,2] ~ factor(tss_rna)))$r.squared

cat("RNA after ComBat:\n")
cat("R² PC1:", round(r2_pc1, 3), "\n")
cat("R² PC2:", round(r2_pc2, 3), "\n")

# Before was:
cat("\nBefore ComBat: PC1=0.235, PC2=0.166\n")
```
```{r batch-after-combat-meth}
tss_meth <- substr(colnames(meth_M), 6, 7)
pca_meth <- prcomp(t(meth_M[order(apply(meth_M, 1, var),
                                    decreasing=TRUE)[1:1000], ]), scale.=TRUE)
r2_meth_pc1 <- summary(lm(pca_meth$x[,1] ~ factor(tss_meth)))$r.squared
r2_meth_pc2 <- summary(lm(pca_meth$x[,2] ~ factor(tss_meth)))$r.squared

cat("Methylation after ComBat:\n")
cat("R² PC1:", round(r2_meth_pc1, 3), "\n")
cat("R² PC2:", round(r2_meth_pc2, 3), "\n")
cat("\nBefore ComBat: PC1=0.167, PC2=0.118\n")
```







```{r clinical-labels}
clin_final <- readRDS("clinical.rds")

# Binary 3-year label — used for UMAP coloring and EDA only
# Cox regression does not use this label
clin_final$label <- NA
clin_final$label[clin_final$vital_status == "Dead" &
                 clin_final$days_to_death <= 1095] <- 0
clin_final$label[clin_final$days_to_last_follow_up > 1095] <- 1

saveRDS(clin_final, "clin_final.rds")
cat("Label distribution (0=died<3yr, 1=survived>3yr):\n")
print(table(clin_final$label, useNA = "always"))
```



## 4c. UMAP — RNA-seq colored by survival label
```{r umap-rna}
library(umap)
label_lookup <- setNames(clin_final$label, clin_final$submitter_id)
top_var      <- order(apply(rna_final, 1, var), decreasing = TRUE)[1:2000]
um_rna       <- umap(t(rna_final[top_var, ]))

ggplot(data.frame(
    UMAP1    = um_rna$layout[,1],
    UMAP2    = um_rna$layout[,2],
    survival = label_lookup[substr(colnames(rna_final), 1, 12)]),
  aes(UMAP1, UMAP2, color = factor(survival))) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_color_manual(values   = c("0" = "#C0392B", "1" = "#2E4A7A"),
                     labels   = c("0" = "Did not survive", "1" = "Survived"),
                     na.value = "grey80") +
  theme_minimal() +
  labs(title = "RNA-seq UMAP — survival label", color = "Survival")
```


## 4d. UMAP — Methylation colored by survival label

```{r umap-meth}
top_cpg  <- order(apply(meth_final, 1, var), decreasing = TRUE)[1:1000]
um_meth  <- umap(t(meth_final[top_cpg, ]))

ggplot(data.frame(
    UMAP1    = um_meth$layout[, 1],
    UMAP2    = um_meth$layout[, 2],
    survival = label_lookup[substr(colnames(meth_final), 1, 12)]),
  aes(UMAP1, UMAP2, color = factor(survival))) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_color_manual(values = c("0" = "#C0392B", "1" = "#2E4A7A"),
                     labels = c("0" = "Did not survive", "1" = "Survived"),
                     na.value = "grey80") +
  theme_minimal() +
  labs(title = "Methylation UMAP — survival label", color = "Survival")
```



## 4e. OncoPrint — top 20 mutated genes

```{r oncoprint}
library(maftools)

maf_obj <- read.maf(maf = mut_data)

grDevices::png("oncoprint.png", width = 1400, height = 900, res = 120)
oncoplot(maf = maf_obj, top = 20, fontSize = 12)
title("Top 20 mutated genes — CRC cohort")
grDevices::dev.off()

# Also display inline
oncoplot(maf = maf_obj, top = 20)
```


## 4f. Tumor Mutation Burden (TMB) distribution

```{r tmb-histogram}
# TMB distribution is bimodal in CRC (MSI-H hypermutated vs. MSS)
tmb <- maf_obj@variants.per.sample

ggplot(tmb, aes(x = Variants)) +
  geom_histogram(bins = 50, fill = "#2E4A7A", color = "white", alpha = 0.8) +
  scale_x_log10() +
  theme_minimal() +
  labs(
    title = "Tumor Mutation Burden — CRC cohort",
    x     = "Number of mutations per patient (log scale)",
    y     = "Number of patients"
  )
```

# ══════════════════════════════════════════════════════════════
# SECTION 5 — CROSS-MODAL PATIENT ALIGNMENT
# Take the intersection of patients present in all three omics
# layers. After alignment, all matrices must have the exact
# same column order (patients) for MOFA2 to work correctly.
# ══════════════════════════════════════════════════════════════

```{r align-patients}
# ── Verify current dimensions before alignment ──
cat("RNA final:  ", dim(rna_final),  "\n")
cat("Meth final: ", dim(meth_final), "\n")
cat("Mut final:  ", dim(mut_final),  "\n")

# ── Confirm column order is identical across layers ──
cat("RNA vs Meth aligned:", all(colnames(rna_final) == colnames(meth_final)), "\n")
cat("RNA vs Mut aligned: ", all(colnames(rna_final) == colnames(mut_final)),  "\n")
```


# ══════════════════════════════════════════════════════════════
# SECTION 6 — MOFA2 MULTI-OMICS INTEGRATION
# MOFA2 (Multi-Omics Factor Analysis v2) learns a low-dimensional
# latent factor space shared across omics layers. Each factor
# captures coordinated variation across modalities and is
# interpretable via per-layer feature weights.
#
# Key choices made here:
#   - 5,000 top-variance features from RNA + methylation
#   - All 100 mutation genes (already filtered)
#   - 15 latent factors (chosen to capture major sources of variation)
#   - Gaussian likelihood for RNA + methylation (continuous)
#   - Bernoulli likelihood for mutations (binary 0/1)
# ══════════════════════════════════════════════════════════════

## 6a. Prepare MOFA2 input

```{r mofa-input}
# ── Select top variable features per continuous layer ──
# High-variance features carry the most signal; low-variance features
# add noise and slow down training considerably.
top_rna_idx  <- order(apply(rna_final,  1, var), decreasing = TRUE)[1:5000]
top_meth_idx <- order(apply(meth_final, 1, var), decreasing = TRUE)[1:5000]

# ── Pack into named list — one matrix per modality ──
# MOFA2 expects features × samples (rows = features, cols = patients)
data_mofa <- list(
  RNA         = rna_final[top_rna_idx,  ],
  Methylation = meth_final[top_meth_idx, ],
  Mutations   = mut_final
)

saveRDS(data_mofa, "data_mofa.rds")

cat("MOFA2 input ready:\n")
cat("RNA:        ", dim(data_mofa$RNA),         "\n")
cat("Methylation:", dim(data_mofa$Methylation), "\n")
cat("Mutations:  ", dim(data_mofa$Mutations),   "\n")
```
## 6b. Create MOFA object

```{r mofa-create}
library(MOFA2)

data_mofa <- readRDS("data_mofa.rds")

mofa <- create_mofa(data_mofa)
cat("MOFA object created\n")
print(mofa)
```

```{r mofa-options}
model_opts                <- get_default_model_options(mofa)
model_opts$num_factors    <- 15
model_opts$likelihoods    <- c(RNA         = "gaussian",
                                Methylation = "gaussian",
                                Mutations   = "bernoulli")
train_opts                <- get_default_training_options(mofa)
train_opts$seed           <- 42

mofa <- prepare_mofa(mofa,
                     model_options    = model_opts,
                     training_options = train_opts)
cat("MOFA2 ready to train\n")
```
```{r mofa-run, eval=FALSE}
# Run once only — takes 30-60 minutes
# eval=FALSE prevents re-running during knit
mofa <- run_mofa(mofa, use_basilisk = TRUE)
saveRDS(mofa, "mofa_model.rds")
cat("MOFA2 trained and saved\n")
```





## 6d. Run MOFA2 training



```{r mofa-reload}
# ── After training: reload from disk ──
mofa <- readRDS("mofa_model.rds")
```


## 6e. Variance explained per factor

```{r mofa-variance}
# ── How much variance does each factor explain in each layer? ──
# Factors that explain variance in multiple layers are the most
# biologically informative (true multi-omics signal).
plot_variance_explained(mofa, max_r2 = 15)

# ── Total variance captured per layer ──
p <- plot_variance_explained(mofa, plot_total = TRUE)
p[[2]]   # extract the total variance bar chart from the returned list
```

```{r extract-factors}
factors <- get_factors(mofa, factors = "all")[[1]]
cat("Factor matrix dimensions:", dim(factors), "\n")
```


```{r silhouette-scores}
library(cluster)

sil_scores <- sapply(2:6, function(k) {
  set.seed(42)
  km <- kmeans(factors, centers = k, nstart = 25)
  mean(silhouette(km$cluster, dist(factors))[, 3])
})

best_k <- (2:6)[which.max(sil_scores)]
cat("Best k:", best_k, "\n")

sil_df <- data.frame(k = 2:6, silhouette = sil_scores)
ggplot(sil_df, aes(k, silhouette)) +
  geom_line(color = "#2E4A7A") +
  geom_point(size = 3, color = "#2E4A7A") +
  geom_vline(xintercept = best_k, linetype = "dashed", color = "#C0392B") +
  theme_minimal() +
  labs(title = "Silhouette Score by Number of Clusters",
       x = "Number of clusters (k)", y = "Mean silhouette width")

```




```{r kmeans-final}
set.seed(42)

km_final       <- kmeans(factors, centers = best_k, nstart = 25)
subtype_labels <- km_final$cluster

cat("Subtype distribution:\n")
print(table(subtype_labels))
```





```{r umap-subtypes}

library(umap)

umap_out <- umap(factors)
df_umap  <- data.frame(
  UMAP1   = umap_out$layout[, 1],
  UMAP2   = umap_out$layout[, 2],
  Subtype = factor(subtype_labels)
)

ggplot(df_umap, aes(UMAP1, UMAP2, color = Subtype)) +
  geom_point(size = 1.5, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Molecular Subtypes — UMAP of MOFA2 Factors",
       color = "Subtype")
```



```{r factor-survival}
library(survival)

# Align factors to clinical
clin_m <- clin_final[match(substr(rownames(factors), 1, 12),
                            clin_final$submitter_id), ]
time   <- ifelse(clin_m$vital_status == "Dead",
                 clin_m$days_to_death, clin_m$days_to_last_follow_up)
event  <- ifelse(clin_m$vital_status == "Dead", 1, 0)

# Log-rank p-value per factor
pval_df <- data.frame(
  Factor  = paste0("Factor", 1:15),
  p_value = sapply(1:15, function(i) {
    d  <- data.frame(time, event, g = ifelse(factors[,i] > median(factors[,i]), "High", "Low"))
    d  <- d[!is.na(d$time) & d$time > 0, ]
    lr <- tryCatch(survdiff(Surv(time, event) ~ g, data = d), error = function(e) NULL)
    if (is.null(lr)) return(NA)
    1 - pchisq(lr$chisq, df = 1)
  })
) |> arrange(p_value)

print(pval_df)

ggplot(pval_df, aes(x = reorder(Factor, p_value), y = p_value)) +
  geom_bar(stat = "identity", fill = "#2E4A7A", alpha = 0.8) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "#C0392B") +
  coord_flip() + theme_minimal() +
  labs(title = "Survival Association per MOFA2 Factor",
       x = "Factor", y = "Log-rank p-value") +
  annotate("text", x = 1, y = 0.06, label = "p = 0.05", color = "#C0392B", size = 3)
```
```{r check-weights}
# Factor 12 RNA — weight distribution
weights_f12_rna <- get_weights(mofa, views = "RNA",       factors = 12)[[1]]
weights_f12_mut <- get_weights(mofa, views = "Mutations", factors = 12)[[1]]
weights_f6_rna  <- get_weights(mofa, views = "RNA",       factors = 6)[[1]]

cat("Factor 12 RNA — top 10 weights:\n")
print(round(sort(abs(weights_f12_rna[,1]), decreasing=TRUE)[1:10], 4))

cat("\nFactor 12 Mutations — top 10 weights:\n")
print(round(sort(abs(weights_f12_mut[,1]), decreasing=TRUE)[1:10], 4))

cat("\nFactor 6 RNA — top 10 weights:\n")
print(round(sort(abs(weights_f6_rna[,1]), decreasing=TRUE)[1:10], 4))
```


```{r export-ml-features}
# Top 25 genes from Factor 12 and Factor 6
get_top <- function(factor_num, n = 25) {
  w <- get_weights(mofa, views = "RNA", factors = factor_num)[[1]]
  data.frame(gene = rownames(w), weight = w[,1]) |>
    arrange(desc(abs(weight))) |> head(n)
}

top_genes <- rbind(get_top(12), get_top(6)) |>
             distinct(gene, .keep_all = TRUE)

cat("Total unique genes:", nrow(top_genes), "\n")
write.csv(top_genes, "top_genes_Factor6.csv", row.names = FALSE)

# Subset RNA and save
common_pts      <- intersect(substr(colnames(rna_final), 1, 12),
                              clin_final$submitter_id)
rna_sub         <- rna_final[rownames(rna_final) %in% top_genes$gene,
                               substr(colnames(rna_final), 1, 12) %in% common_pts]
ml_df           <- as.data.frame(t(rna_sub))
colnames(ml_df) <- make.names(colnames(ml_df))

cat("Patients:", nrow(ml_df), "| Features:", ncol(ml_df), "\n")
write.csv(ml_df, "ml_f6_features.csv", row.names = TRUE)
cat("Saved\n")

```

```{r}
meta <- read.table(
  gzfile("GSE132465/GSE132465_GEO_processed_CRC_10X_cell_annotation.txt.gz"),
  header = TRUE, row.names = 1, sep = "\t"
)
cat("Total cells:", nrow(meta), "\n")
cat("Sample types:\n")
print(table(meta$Class))   # tumor vs normal
cat("Cell types:\n")
print(table(meta$Cell_type))
```

```{r}

library(data.table)
library(Seurat)

set.seed(42)

# keep only tumor cells, exclude mast cells (only 3 tumor cells — not useful)
tumor_meta <- meta[meta$Class == "Tumor" & meta$Cell_type != "Mast cells", ]

# pick 500 cells per cell type
sampled_cells <- do.call(rbind, lapply(
  unique(tumor_meta$Cell_type), function(ct) {
    cells <- rownames(tumor_meta[tumor_meta$Cell_type == ct, ])
    n     <- min(500, length(cells))
    data.frame(barcode = sample(cells, n), Cell_type = ct)
  }
))

cat("Selected cells:\n")
print(table(sampled_cells$Cell_type))
cat("Total:", nrow(sampled_cells), "\n")

# load only those columns from the big matrix
header    <- fread("GSE132465/GSE132465_GEO_processed_CRC_10X_natural_log_TPM_matrix.txt.gz", nrows = 0)
all_cols  <- names(header)
keep_cols <- c(all_cols[1], sampled_cells$barcode[sampled_cells$barcode %in% all_cols])

cat("Loading matrix...\n")
mat_small <- fread("GSE132465/GSE132465_GEO_processed_CRC_10X_natural_log_TPM_matrix.txt.gz", select = keep_cols)
mat <- as.matrix(mat_small[, -1])
rownames(mat) <- mat_small[[1]]

meta_small <- meta[colnames(mat), ]
sc <- CreateSeuratObject(counts = mat, meta.data = meta_small)
cat("Done:", ncol(sc), "cells x", nrow(sc), "genes\n")

saveRDS(sc, "sc_small.rds")
```
```{r scrna-process}
library(Seurat)

sc <- readRDS("sc_small.rds")

# QC — remove dead cells and doublets
sc[["percent.mt"]] <- PercentageFeatureSet(sc, pattern = "^MT-")
sc <- subset(sc, subset = nFeature_RNA > 200 &
                          nFeature_RNA < 6000 &
                          percent.mt   < 25)
cat("After QC:", ncol(sc), "cells\n")

# Normalise — required for meaningful expression comparison
sc <- NormalizeData(sc)

saveRDS(sc, "sc_processed.rds")
cat("Saved: sc_processed.rds\n")
```


```{r convert-and-overlap}
library(org.Hs.eg.db)

# Load top 50 Factor 6 and 12 genes from saved file
top_genes_f6 <- read.csv("top_genes_Factor6.csv")$gene

# Strip version numbers from ENSEMBL IDs
ensembl_clean <- gsub("\\..*", "", top_genes_f6)

# Convert ENSEMBL to gene symbols
gene_symbols <- mapIds(
  org.Hs.eg.db,
  keys      = ensembl_clean,
  column    = "SYMBOL",
  keytype   = "ENSEMBL",
  multiVals = "first"
)

# Remove failed conversions
gene_symbols <- gene_symbols[!is.na(gene_symbols)]
cat("Successfully converted:", length(gene_symbols), "genes\n")
print(gene_symbols)

# Check overlap with single-cell dataset
overlap <- intersect(gene_symbols, rownames(sc))
cat("Genes found in scRNA-seq:", length(overlap), "out of", length(gene_symbols), "\n")
```

```{r factor6-heatmap}
library(Seurat)
library(ggplot2)
library(tidyr)
library(org.Hs.eg.db)

sc <- readRDS("sc_processed.rds")
Idents(sc) <- "Cell_type"

# Convert ENSEMBL to symbols
genes   <- gsub("\\..*", "", read.csv("top_genes_Factor6.csv")$gene)
symbols <- mapIds(org.Hs.eg.db, keys = genes, column = "SYMBOL",
                  keytype = "ENSEMBL", multiVals = "first")
overlap <- intersect(symbols[!is.na(symbols)], rownames(sc))
cat("Genes found in scRNA:", length(overlap), "\n")

# Average expression per cell type
avg_long <- AverageExpression(sc, features = overlap,
                               group.by = "Cell_type")$RNA |>
            as.data.frame() |>
            tibble::rownames_to_column("gene") |>
            pivot_longer(-gene, names_to = "cell_type",
                         values_to = "expression")

# Heatmap — saved as large PNG only
p <- ggplot(avg_long, aes(x = cell_type, y = gene, fill = expression)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_gradient(low = "lightgrey", high = "#C0392B") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
        axis.text.y = element_text(size = 9, face = "bold"),
        panel.grid  = element_blank(),
        plot.title  = element_text(face = "bold", size = 15)) +
  labs(title = "Factor 6 & 12 Genes — Expression per Cell Type",
       x = "Cell Type", y = NULL, fill = "Avg Expression")

ggsave("factor6_heatmap.png", plot = p,
       width = 12, height = 18, dpi = 200)
cat("Saved: factor6_heatmap.png\n")
```


```{r cox-regression}
library(survival)
library(survminer)

# Load ML features
ml_df <- read.csv("ml_f6_features.csv", row.names = 1)

# Match clinical data directly
sample_ids   <- substr(rownames(ml_df), 1, 12)
clin_matched <- clin_final[match(sample_ids, clin_final$submitter_id), ]

# Use existing clinical columns directly
ml_df$time  <- ifelse(clin_matched$vital_status == "Dead",
                       clin_matched$days_to_death,
                       clin_matched$days_to_last_follow_up)
ml_df$event <- ifelse(clin_matched$vital_status == "Dead", 1, 0)
ml_df$label <- NULL

ml_df <- ml_df[!is.na(ml_df$time) & ml_df$time > 0, ]
cat("Patients in Cox:", nrow(ml_df), "\n")

# Fit Cox
cox_fit <- coxph(Surv(time, event) ~ ., data = ml_df, x = TRUE)
cat("C-index:", round(cox_fit$concordance["concordance"], 3), "\n")
summary(cox_fit)
```

```{r event-summary}
sample_ids   <- substr(rownames(ml_df), 1, 12)
clin_matched <- clin_final[match(sample_ids, clin_final$submitter_id), ]

cat("Total patients in Cox:    ", nrow(ml_df), "\n")
cat("Total deaths (events):    ", sum(ml_df$event == 1), "\n")
cat("Censored (alive/unknown): ", sum(ml_df$event == 0), "\n")
cat("Median follow-up (days):  ", round(median(ml_df$time), 0), "\n")
cat("Max follow-up (days):     ", max(ml_df$time), "\n")
```
```{r cox-km}
# Use linear predictors directly — no predict needed
pred <- cox_fit$linear.predictors
ml_df$risk_group <- ifelse(pred > median(pred), "High", "Low")

fit_risk <- survfit(Surv(time, event) ~ risk_group, data = ml_df)

ggsurvplot(fit_risk, data = ml_df, pval = TRUE,
           palette      = c("#C0392B", "#2E4A7A"),
           title        = "KM — High vs Low Cox Risk Score",
           xlab         = "Days",
           ylab         = "Survival Probability",
           legend.title = "Risk Group",
           ggtheme      = theme_minimal())
```


```{r top-cox-genes-plot}
cox_summary <- as.data.frame(summary(cox_fit)$coefficients)
cox_summary$gene <- rownames(cox_summary)

top_genes_cox <- cox_summary |>
  arrange(desc(abs(coef))) |>
  dplyr::select(gene, coef, `exp(coef)`, `Pr(>|z|)`) |>
  head(15)

colnames(top_genes_cox) <- c("gene", "coef", "hazard_ratio", "p_value")

# Direction label
top_genes_cox$direction <- ifelse(top_genes_cox$coef > 0, "Risk", "Protective")

ggplot(top_genes_cox, aes(x = reorder(gene, abs(coef)),
                           y = coef,
                           fill = direction)) +
  geom_bar(stat = "identity", width = 0.7, alpha = 0.85) +
  scale_fill_manual(values = c("Risk" = "#C0392B", "Protective" = "#2E4A7A")) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 15 Genes by Factor 12&6 Driving Cox Survival Model",
       x     = "Gene",
       y     = "Cox Coefficient",
       fill  = "Direction") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  theme(plot.title = element_text(face = "bold", size = 13))
```
```{r gene-annotation}
library(org.Hs.eg.db)
library(dplyr)

# Get top 15 Cox genes
cox_summary      <- as.data.frame(summary(cox_fit)$coefficients)
cox_summary$gene <- rownames(cox_summary)

top15 <- cox_summary |>
  arrange(desc(abs(coef))) |>
  head(15)

# Strip version numbers
ensembl_clean <- gsub("\\..*", "", top15$gene)

# Convert to symbols and full names
symbols  <- mapIds(org.Hs.eg.db, keys = ensembl_clean,
                   column = "SYMBOL",   keytype = "ENSEMBL", multiVals = "first")
genename <- mapIds(org.Hs.eg.db, keys = ensembl_clean,
                   column = "GENENAME", keytype = "ENSEMBL", multiVals = "first")

# Build table
annot_df <- data.frame(
  symbol       = symbols,
  full_name    = genename,
  hazard_ratio = round(top15$`exp(coef)`, 3),
  direction    = ifelse(top15$coef > 0, "Risk", "Protective")
)

print(annot_df)
write.csv(annot_df, "top15_annotated_genes.csv", row.names = FALSE)
cat("Saved: top15_annotated_genes.csv\n")
```


```{r gprofiler-simple}
library(gprofiler2)

# Your annotated gene symbols
gene_list <- annot_df$symbol[!is.na(annot_df$symbol)]
cat("Genes for enrichment:", paste(gene_list, collapse=", "), "\n")

# Run enrichment
gost_res <- gost(query = gene_list,
                 organism = "hsapiens",
                 sources  = c("GO:BP", "KEGG", "REAC"),
                 correction_method = "fdr")

# Top pathways
top_paths <- gost_res$result |>
  dplyr::arrange(p_value) |>
  dplyr::select(source, term_name, p_value, intersection_size) |>
  head(15)

print(top_paths)
```
```{r gprofiler-plot}
library(gprofiler2)
library(ggplot2)

# Load annotated genes
annot_df  <- read.csv("top15_annotated_genes.csv")
gene_list <- annot_df$symbol[!is.na(annot_df$symbol)]

# Run enrichment
gost_res <- gost(query             = gene_list,
                 organism          = "hsapiens",
                 sources           = c("GO:BP", "KEGG", "REAC"),
                 correction_method = "fdr")

# Top 15 pathways
top_paths <- gost_res$result |>
  dplyr::arrange(p_value) |>
  dplyr::select(source, term_name, p_value, intersection_size) |>
  head(15)

# Plot as bar chart — saves as image
ggplot(top_paths, aes(x = reorder(term_name, -log10(p_value)),
                       y = -log10(p_value),
                       fill = source)) +
  geom_bar(stat = "identity", alpha = 0.85) +
  coord_flip() +
  scale_fill_manual(values = c("GO:BP" = "#2E4A7A",
                                "KEGG"  = "#C0392B",
                                "REAC"  = "#27AE60")) +
  theme_minimal() +
  theme(axis.text.y  = element_text(size = 9),
        plot.title   = element_text(face = "bold", size = 13),
        legend.title = element_text(size = 10)) +
  labs(title = "Pathway Enrichment — Top Cox Survival Genes",
       x     = "Pathway",
       y     = "-log10(adjusted p-value)",
       fill  = "Database")

ggsave("gprofiler_barplot.png", width = 12, height = 8, dpi = 200)
cat("Saved: gprofiler_barplot.png\n")
```